---
title: "CARAMeL visualization"
author: "Carolina H. Chung"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_notebook: 
    code_folding: hide
---

This is an R Markdown Notebook for generating figure components for the CARAMeL manuscript. 

# Set up environment

Install required packages.

```{r install_packages, results='hide', message=F, warning=F, error=F}

# Define packages to install
pkg.list = c('svDialogs', 'dplyr', 'readxl', 'RColorBrewer', 'pheatmap')

# Define packages not already installed
pkg.install <- pkg.list[!(pkg.list %in% installed.packages()[, 'Package'])]

# Install pheatmap from GitHub 
if ('pheatmap' %in% pkg.install) {
  if (!require('devtools')) {
    install.packages('devtools', dependencies=T)
    library(devtools)
  }
  install_github('raivokolde/pheatmap')
  pkg.install <- pkg.install[!(pkg.install %in% 'pheatmap')]
}

# Install other uninstalled packages
if (length(pkg.install) > 0) {
  install.packages(pkg.install)
}

```

Load installed packages. 

```{r load packages, include=F}

library(svDialogs)    # for user-defined inputs
library(dplyr)        # for pipe-based processing
library(readxl)       # for reading Excel files
library(RColorBrewer) # for custom color palettes
library(pheatmap)     # for generating heatmaps
  
```

# Set user-defined inputs

Define user-defined inputs.

```{r user_inputs}

# Color palette
col <- c(brewer.pal(9, 'Set1'), brewer.pal(3, 'Accent'))

# Save plots?
save.plot <- dlgInput('Save heatmaps? (T/F)', Sys.info()['F'])$res %>% 
  as.logical(.)

```

# Load data 

Load data to visualize. Specifically: 

- *E. coli* drug-media interactions (Biolog)
- *E. coli* CARAMeL model predictions
- *E. coli* cell-specific fluxes (populationFBA)

```{r load data}

# Create dataset variable
dataset <- list()

# EC (Biolog)
dataset$EC$bio <- read_excel('results/plot_data.xlsx', sheet='ecoli_biolog')

# EC (predictions)
dataset$EC$pred <- read_excel('results/plot_data.xlsx', sheet='pred2_heatmap')
dataset$EC$pred3 <- read_excel('results/plot_data.xlsx', sheet='pred3_heatmap')
dataset$EC$predp <- read_excel('results/plot_data.xlsx', sheet='popFBA_heatmap')

# popFBA (fluxes)
dataset$EC$fluxp <- read_excel('results/plot_data.xlsx', sheet='popFBA_flux')
  
```

# Generate plots

Biolog heatmap. 

```{r E. coli interactions (Biolog), warning = F}

# Re-organize data
dataset$EC$bio$Media <- dataset$EC$bio$Media %>% factor(., levels=unique(.))
y <- dataset$EC$bio[, -c(1:2)]; rownames(y) <- dataset$EC$bio$Media

# Define metadata
gx <- data.frame(Simulated=dataset$EC$bio$Simulatable %>% 
                   factor(., levels=unique(.)))
rownames(gx) <- rownames(y)
gy <- data.frame(Subset=c(rep('train', 4), rep('test', 2)) %>% 
                  factor(., levels=unique(.)), 
                 Target=c(rep('Cell wall', 2), 
                          rep('Protein synthesis, 30S', 2), 
                          rep('Multiple mechanisms', 2)) %>% 
                  factor(., levels=unique(.)))
rownames(gy) <- colnames(y)

# Use custom color palette (asymmetric)
col.pal <- colorRampPalette(colors=c('snow2', 'orchid3'), space='Lab')(100)

# Define annotation colors
col.annot <- list(Simulated=c(Yes='limegreen', No='gray'), 
                  Subset=c(train='lightgoldenrod', test='green4'), 
                  Target=c(`Cell wall`='darkorange', 
                           `Protein synthesis, 30S`='turquoise3',
                           `Multiple mechanisms`='sienna'))
names(col.annot$Simulated) <- levels(gx$Simulated)
names(col.annot$Subset) <- levels(gy$Subset)
names(col.annot$Target) <- levels(gy$Target)
gz <- data.frame(Target=gy$Target)
rownames(gz) <- rownames(gy)

# Visualize plot
p1 <- pheatmap(t(y), cluster_rows=F, cluster_cols=F, gaps_row=4, gaps_col=96,
         legend_breaks=c(min(y), 0.5, 1, 1.5, max(y)-0.05), border_color=T,
         legend_labels=c('0','0.5','1','1.5','Respiration'),
         annotation_row=gz, annotation_col=gx, annotation_colors=col.annot,
         color=col.pal, angle_col='90', show_colnames=F, cellheight=15,
         cellwidth=2, main='Drug-media interactions')
p1

# Save plot (if prompted)
if (save.plot) {
  tiff('results/R_plots/biolog_heatmap.tiff', units='in', 
       width=7, height=2.5, res=300)
  print({
    p1
  })
  dev.off()
}

```

Heatmap of 2-way predictions. 

```{r E. coli predictions (2-way), warning = F}

# Re-organize data
y <- dataset$EC$pred[, -1]; rownames(y) <- dataset$EC$pred$Pair

# Use RedBlue color palette (asymmetric)
n <- round(100*(max(y) - min(y)))
x <- n + round(100*min(y))
col.neg <- colorRampPalette(colors=c('royalblue3', 'white'), 
                            space='Lab')(abs(round(100*min(y))))
col.pos <- colorRampPalette(colors=c('white', 'red4'), 
                            space='Lab')(abs(round(100*max(y))))
col.pal <- c(col.neg, col.pos)

# Visualize plot
p2 <- pheatmap(y, cluster_rows=F, cluster_cols=F, gaps_col=c(57, 114),
         clustering_distance_rows='manhattan', color=col.pal, 
         legend_breaks=c(min(y)+0.03, 0, max(y)-0.03), border_color=NA, 
         legend_labels=c('Synergy','Neutral','Antagonism'), 
         show_rownames=F, show_colnames=F, cellheight=2, cellwidth=2)
p2

# Save plot (if prompted)
if (save.plot) {
  tiff('results/R_plots/pred2_heatmap.tiff', units='in', 
       width=6.5, height=2.4, res=300)
  print({
    p2
  })
  dev.off()
}

```

Heatmap of 3-way predictions. 

```{r E. coli predictions (3-way), warning = F}

# Re-organize data
y <- dataset$EC$pred3[, -1]
rownames(y) <- dataset$EC$pred3$Pair

# Use RedBlue color palette (asymmetric)
n <- round(100*(max(y) - min(y)))
x <- n + round(100*min(y))
col.neg <- colorRampPalette(colors=c('royalblue3', 'white'), 
                            space='Lab')(abs(round(100*min(y))))
col.pos <- colorRampPalette(colors=c('white', 'red4'), 
                            space='Lab')(abs(round(100*max(y))))
col.pal <- c(col.neg, col.pos)

# Visualize plot
p2 <- pheatmap(y, cluster_rows=F, cluster_cols=F, 
               gaps_col=c(57, 2*57, 3*57, 4*57, 5*57, 6*57),
         clustering_distance_rows='manhattan', color=col.pal, 
         legend_breaks=c(min(y)+0.03, 0, max(y)-0.03), border_color=NA, 
         legend_labels=c('Synergy','Neutral','Antagonism'), 
         show_rownames=F, show_colnames=F, cellheight=1, cellwidth=1)
p2

# Save plot (if prompted)
if (save.plot) {
  tiff('results/R_plots/pred3_heatmap.tiff', units='in', 
       width=7.5, height=2.5, res=300)
  print({
    p2
  })
  dev.off()
}

```

Heatmap of cell-specific predictions (popFBA). 

```{r E. coli predictions (popFBA), warning = F}

# Re-organize data
x <- dataset$EC$fluxp %>% 
  relocate(colnames(.)[.[1, ]==2], .after=last_col())
y <- dataset$EC$predp[, -1]
rownames(y) <- dataset$EC$predp$Pair
y <- y[, colnames(y) %in% colnames(x)] %>%  .[, colnames(x)[-c(1, 2)]] %>% 
  rbind(x[1, -c(1, 2)], .) %>% 
  relocate(colnames(.)[.[1, ]==2], .after=last_col())
z1 <- y[-1, colnames(y)[y[1, ]==1]]
z2 <- y[-1, colnames(y)[y[1, ]==2]]
y <- cbind(z1[, 1:50], z2[, 1:50])

# Use RedBlue color palette (asymmetric)
n <- round(100*(max(y) - min(y)))
x <- n + round(100*min(y))
col.neg <- colorRampPalette(colors=c('royalblue3', 'white'), 
                            space='Lab')(abs(round(100*min(y))))
col.pos <- colorRampPalette(colors=c('white', 'red4'), 
                            space='Lab')(abs(round(100*max(y))))
col.pal <- c(col.neg, col.pos)

# Annotations
Cluster <- data.frame(Cluster=as.character(rep(c(1, 2), each=50)), 
                      row.names=colnames(y))
annot.pal <- list(Cluster=c('1'='goldenrod', '2'='darkgreen'))

# Visualize plot
p2 <- pheatmap(y, cluster_rows=F, cluster_cols=F, gaps_col=c(50),
         clustering_distance_rows='manhattan', color=col.pal, 
         legend_breaks=c(min(y)+0.01, 0, max(y)-0.01), border_color=NA, 
         annotation_col=Cluster, annotation_colors=annot.pal, 
         legend_labels=c('Synergy','Neutral','Antagonism'), 
         show_rownames=F, show_colnames=F, cellheight=2, cellwidth=2, 
         main='Predicted interaction scores')
p2

# Save plot (if prompted)
if (save.plot) {
  tiff('results/R_plots/popFBA_heatmap.tiff', units='in', 
       width=5, height=2, res=300)
  print({
    p2
  })
  dev.off()
}

```

Heatmap of cell-specific fluxes (popFBA). 

```{r popFBA flux, warning = F}

# Re-organize data
x <- dataset$EC$fluxp %>% select(rxn, subSystem, colnames(y))
# y <- x[-1, -c(1, 2)]
y <- x[-1, -c(1, 2)] %>% t(.) %>% scale(.) %>% t(.) %>% 
  as.data.frame(., row.names=make.unique(x$rxn[-1])) %>% 
  arrange()
# y[y > 2] <- 2; y[y < -2] <- -2
rownames(y) <- x$rxn[-1] %>% make.unique(.)
colnames(y) <- colnames(x)[-c(1, 2)]

# Define metadata
Pathway <- data.frame(Pathway=x$subSystem[-1], row.names=rownames(y))
Cluster <- data.frame(Cluster=as.character(x[1, -c(1, 2)]), 
                      row.names=colnames(y))

# Use RedBlue color palette 
col.neg <- colorRampPalette(colors=c('royalblue3', 'white'), 
                            space='Lab')(abs(round(100*min(y))))
col.pos <- colorRampPalette(colors=c('white', 'red4'), 
                            space='Lab')(abs(round(100*max(y))))
col.pal <- c(col.neg, col.pos)

# Use custom annotation palette 
col <- rev(brewer.pal(dim(unique(Pathway))[1], 'Set3'))
names(col) <- unique(Pathway$Pathway)
annot.pal <- list(Pathway=col, Cluster=c('1'='goldenrod', '2'='darkgreen'))

# Visualize plot
p2 <- pheatmap(y, scale='none', show_rownames=F, show_colnames=F, 
         cluster_rows=T, clustering_distance_rows='manhattan',
         cluster_cols=F, annotation_row=Pathway, annotation_col=Cluster, 
         color=col.pal, annotation_colors=annot.pal, borker_color='gray', 
         cellwidth=2, cellheight=2, gaps_col=match(2, Cluster$Cluster)-1, 
         main='Flux through metabolic reactions', treeheight_row = 0)
p2

# Save plot (if prompted)
if (save.plot) {
  tiff('results/R_plots/popFBA_flux.tiff', units='in', 
       width=7, height=3, res=300)
  print({
    p2
  })
  dev.off()
}

```
